---
title: "C) Data subsetting using dplry"
author: "Nathan Brouwer"
date: "November 13, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Make some subset without certain columns and certain rows

## Preliminaries

```{r}
library(here)
library(dplyr)
```

## Load data

Load data cleaned in previous script
```{r}
file. <- "skibiel_mammalsmilk.csv"

full.path <- here::here("data", #folder
                        file.)  #file name

dat.selected <- read.csv(file = full.path)
```

Check

```{r}
dim(dat.selected)
```


## Remove columns

The N column is sample size and isn't needed.  For many analyses we'll fous on fat, so we'll dorp the other columns

```{r}
dat.fat <- dat.selected %>% dplyr::select(-N,
                                               -prot,
                                               -sugar,
                                               -energy)
```


We'll save this sub-dataset for easy access

```{r}
file. <- "skibiel_fat.csv"

full.path <- here::here("data", #folder
                        file.)  #file name

write.csv(dat.fat,file = full.path,
          row.names =F)
```


## Filter by rows

For some analyses we'll focus on just primates and close-ish relatives.


```{r}
primates.and.friends <- c("Rodentia","Primates","Lagomorpha")

primate.fat <-  dat.fat %>% filter(ord %in% primates.and.friends)
```



```{r}
file. <- "skibiel_primate_fat.csv"

full.path <- here::here("data", #folder
                        file.)  #file name

write.csv(primate.fat,file = full.path,
          row.names =F)
```


```{r}

```

