---
title: "d) Data Exploration for Regression Analyses"
author: "Nathan Brouwer | brouwern@gmail.com | @lowbrowR"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, strip.white = FALSE, cache = TRUE)
```



## Introduction

The data collated by Skibiel et al contain information on several components of milk, including fat, protein, sugar, and total energetic content ("energy").  Here we'll look at multple y variables, though for most of the subsequent tutorials we'll focus on % of fat in milk

The data also contains several "continous predictors" / covariates (x variables).  These include

* size of mom = "mass.fem" 
* duration of pregnancy/gestation = "gest.mon"    
* how long young fed = "lacat.mo" ("number of lactation months")   
* how big the babies are = "mass.litter"  

THere are also Several "categorical predictors"

* "diet" = meat, plants, everything
* "arid" = desert or other; theory predicts arid environments impact evolution of milk
* "biome" = aquatic or terrestrial (includes bats as terrestrial)

Finally, there are columns containing taxonomic information 

* order = order
* fam = family
* spp = species


## Preliminaries

### Load packages

```{r}
library(ggplot2)
library(cowplot)
library(ggpubr)
library(arm)
library(dplyr)
library(GGally)
```



### Loading the mammalsmilk package

If you haven't already, download the mammalsmilk package from GitHub (note the "S" between mammal and milk).  This is commented out in the code below - only run the code if you have never downloaded it before, or haven't recently (in case its been updated)

```{r}
# install_github("brouwern/mammalsmilk")
```

You can then load the package with library()
```{r}
library(mammalsmilk)
```



## Load data


```{r}
data("milk")
```



## Data exploration

Look at the data
    
```{r}
# Hom much data?
dim(milk)

#focal columns
##(use negative indexing to drop some)
summary(milk[,-c(2,3,8,9)])
```



### Focal families in dataset

This code is dense but let's us look at the most common families

Tabulate number of families
```{r}
table.fam <- table(milk$fam)
```

Determine order in data set (not taxonomic order)
```{r}
numeric.order. <- order(table.fam,
                        decreasing = T)
```

Largest families, in order
```{r}
table.fam <- data.frame(table.fam[numeric.order.][1:13])

names(table.fam) <- c("fam","freq")

table.fam
```

Plot as barplot; some added gplot code to rotate the axis lables
```{r}
ggbarplot(data = table.fam,
          y = "freq",
          x = "fam") +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, 
                                   vjust = 0.5))
```



Tabulate number of orders
```{r}
table.ord <- table(milk$ord)
```

Determine order (numeric order, not taxonomic order)
```{r}
numeric.order. <- order(table.ord,
                        decreasing = T)
```

Largest orders, in order
```{r}
table.ord <- data.frame(table.ord[numeric.order.][1:9])

names(table.ord) <- c("ord","freq")

table.ord
```


* Artiodactyla = 23; Even-toed ungulate
* Carnivora = 23
* Primates = 22
* Rodentia = 17
* Chiroptera = 10; bats
* Diprotodontia = 10; marsupials
* Perrissodactyla = 7; odd-toed ungulates
* Cetacea = 6
* Lagomorpha = 3


```{r}
ggbarplot(data = table.ord,
          y = "freq",
          x = "ord") +
  theme(axis.text.x = element_text(angle = 90, 
                                   hjust = 1, 
                                   vjust = 0.5))
```




### Plotting scatter plots

To condense my plotting code I will save titles and axis labels to R objects.

Note that for ylab. I insert a line break by putting a backslash in from of an n, for "new line"
```{r}
main. <- "Regression data: continous vs continuous"
xlab. <- "Continous x: mass of female"
ylab. <- "Continous y: \nfat contenxt of milk"
```


#### Look at % milk fat vs. mass of female

Note that the plot is dominated an outlier, which the whales!

```{r}
ggscatter(y = "fat",
      x = "mass.fem",
      data = milk,
      main = main.,
      xlab = xlab.,
      ylab = ylab.)
```


A similar graph can be made using the stock ggplot function qplot(), except the variables are not quoted
```{r gg1, eval = F}
qplot(y = fat,       # no " " !
      x = mass.fem,
      data = milk,
      main = main.,
      xlab = xlab.,
      ylab = ylab.) 
```




## Look at  milk fat vs. log(mass of female)

* Logging helps linearize the data
* NOTE: statiticians typically use natural log = ln()
* Biologists, especially when allometry is invovled, typically use base 10 = log10()
* Original authors used log10, but I just realized this, so I'll stick with ln()
* I believe (?) give similar answers, but log10 I think (?) makes biological interpretation easier (???)
* ggpubr requires hard-coded variables and can't do the log transform "on the fly" in the same way as ggplot/qplot

We can use the mutate() function in dplyr to take the log.
```{r}
milk <- milk %>% dplyr::mutate(mass.fem.log = log(mass.fem))
```

Look at transformed data
```{r}
summary(milk$mass.fem.log)
```


Plot transformed data
```{r}
main.new <-"Regression data: continous vs log(continuous)" 

ggscatter(y = "fat",
      x = "mass.fem.log",
      data = milk,
      main = main.new,
      ylab = ylab.,
      xlab = xlab.)
```



## Add "smoother"

Helps visualize trends

```{r}
ggscatter(y = "fat",
      x = "mass.fem.log",
      add = "loess",     #smoother
      data = milk,
      main = main.new,
      ylab = ylab.,
      xlab = xlab.)
```





## Look at % milk fat by a categorical variable (diet)

Color-code diets.  This might take a second to render.

```{r}
ggscatter(y = "fat",
      x = "mass.fem.log",
      color = "diet",
      add = "loess",
      data = milk,
      main = main.new,
      ylab = ylab.,
      xlab = xlab.)
```




## Look at % milk fat by diet and biome using facets

* "facets" are a powerful tool in ggplot
* Turns out ALL of our aquatic spp are carnivores (sorry, no manatees)
* This shows that "aquatic" and "diet" are "Confounded"


```{r}
ggscatter(y = "fat",
      x = "mass.fem.log",
      color = "diet",
      facet.by = "biome", # facet command
      data = milk,
      main = main.new,
      ylab = ylab.,
      xlab = xlab.)
```




## Boxplots

It can be useful to make boxplot just by categorical variables.  For me, this plot highlights that almost all carnivores are aquatic, and they have a wide range of fat contents.

```{r, echo = T}
ggboxplot(data = milk,
          add = "jitter",
          color = "biome",
          y = "fat",
          x ="diet")

```








## Focal data subset: Primates & Relatives

Let's look at just a subset of the data
* Primates
* Rodents (this is why we use mouse models)
* Rabbits

```{r}
data("milk_primates")
```



Compare original data and our working subset

* Original = 130 speices
* Working data = 42

```{r}
# Original data
dim(milk)

#out working subset
dim(milk_primates)
```


## Plot working data:  milk fat ~ female size

Do log transform
```{r}
milk_primates <- milk_primates %>% dplyr::mutate(mass.fem.log = log(mass.fem))
```


Make scatterplot; **ggscatterhist()** allows plots to be put in margin
```{r, echo = T}
ggscatterhist(y = "fat",
      x = "mass.fem.log",
      data = milk_primates,
      margin.plot = "boxplot") 
```



## Add labels

This code is fairly dense.  What I'll do is annotate some selected data points with the species names.

First, make some labels.
```{r, echo = T}
#make column to hold particular spp.names
milk_primates$spp.focal <- ""

#select spp of interest
## This is just accross a span of the data
i.focal <- c(17,18,19,23,30,34,39)

#look at focals
milk_primates$spp.focal[i.focal]

#add spp names to spp.focal column
milk_primates$spp.focal[i.focal] <- as.character(milk_primates$spp[i.focal])

milk_primates$spp.focal <- gsub("(^.*)( )(.*)","\\1",milk_primates$spp.focal)

```

plot with names using some fancier ggplot code.  
```{r}
ggplot(data = milk_primates,
       aes(y = fat,
      x = mass.fem.log)) + 
  geom_point()+
  geom_text(aes(label=spp.focal),
            hjust=1, 
            vjust=0)
```




## Add regression line

* % Fat generally declines as body mass increases
* Grey areas is "confidence band"


```{r, echo = T}
ggscatter(y = "fat",
      x = "mass.fem.log",
      add = "reg.line",   #regression line
      conf.int = TRUE,    #confidence int.
      data = milk_primates) 
```


Color code by
```{r}
ggscatter(y = "fat",
      x = "mass.fem.log",
      color = "ord",
      add = "reg.line",   #regression line
      conf.int = TRUE,    #confidence int.
      data = milk_primates) 
```



## Alternate predictors

We can also look at othe predictor besides the size of the mother, such as how many months are spent gestating.

First, we should log transform them.

```{r}
milk_primates <- milk_primates %>% dplyr::mutate(gest.mo.log = log(gest.month),
    lacat.mo.log = log(lacat.mo),
    mass.litter.log = log(mass.litter))
```



### Duration of gestation

Let's look at a relationship fat ~ log(gest.month)

Compare all data compiled with by order

```{r, echo = T}
#all data combined
no.order <- ggscatter(y = "fat",
      x = "gest.mo.log",
      data = milk_primates,
      #color = "ord",
      add = "reg.line",
      conf.int = TRUE)

#split by order
by.order <- ggscatter(y = "fat",
      x = "gest.mo.log",
      data = milk_primates,
      color = "ord",
      add = "reg.line",
      conf.int = TRUE)

plot_grid(no.order,by.order)
```

% Fat generally declines as gestation duration increases.  So, the longer animal is pregnant, the less fatty the milk is.  However, this trend largely disappears when we plot by order.  That is because the Lagomorph data is vary noisy, resulting in a flat line.  The negative trend seen when looking at all of the data combined is largely drive by the fact that lagomorphs tend to have higher milkfat and shorter gestations, and primates have lower milk fat and longer gestations.  This trend is therefore probably driven mostly by differenes between orders, not due to an underlying relationship with gestation.




## Duration of lactation

Another relationship to consider fat ~ log(lacat.mo)

Compare all data compiled with by order
```{r, echo = T}
no.order <- ggscatter(y = "fat",
      x = "lacat.mo.log",
      data = milk_primates,
      #color = "ord",
      add = "reg.line",
      conf.int = TRUE)

by.order <- ggscatter(y = "fat",
      x = "lacat.mo.log",
      data = milk_primates,
      color = "ord",
      add = "reg.line",
      conf.int = TRUE)

plot_grid(no.order,by.order)
```

Overall, the % Fat generally declines as lactation duration increases.  So, the longer time children are dependent, lower milk fat.

Again, however, there is strong phyolgenetic single.  Within





## Mass of litter

 fat ~ log(mass.litter)

* % Fat generally declines as litter size  increases
* more or bigger babies, less milk fat

```{r, echo = T}
no.order <- ggscatter(y = "fat",
      x = "mass.litter.log",
      data = milk_primates,
      #color = "ord",
      add = "reg.line",
      conf.int = TRUE)

by.order <- ggscatter(y = "fat",
      x = "mass.litter.log",
      data = milk_primates,
      color = "ord",
      add = "reg.line",
      conf.int = TRUE)

plot_grid(no.order,by.order)
```




## Correlations among predictors

ALl of the graphs of all of the predictors looked basically the same.  One reason is that all of the predictors are highly correlated, so they are all basically indicating something similar about the animals.

We can get a sense for these correlations using the ggpairs() fucntion in the GGally package.

```{r}
vars. <- c("fat","gest.mo.log",
           "mass.fem.log","lacat.mo.log"  ,  "mass.litter.log")
GGally::ggpairs(data = milk_primates,
        column = vars.,
        diag =NULL )
```


```{r}

```


