---
title: "f2) Running and plotting Linear Regression "
author: "Nathan Brouwer | brouwern@gmail.com | @lowbrowR"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, strip.white = FALSE, cache = FALSE)
```

## Introduction

This walk through introduces how to run a basic linear regression model and plot the output.  It demonstrates different ways to plot the output, including

* ggplot
* ggpubr::ggscatter
* The gg_lm_plot() function in the mammalsmilk package

In this tutorial, we will examine the relationship between the fat content of the milk of different species as a function of the mass of the mother.  We will focus on primate species and their relatively close relatives (rabbits and rodents - yes, they are our relatives).

## Preliminaries

### Load data 

Install the mammals milk package if needed.
```{r install.git.chunk}
# install_github("brouwern/mammalsmilk")
```

You can then load the package with library()
```{r}
library(mammalsmilk)
```


The example data is milk_primates
```{r}
data("milk_primates")
```


### Load libraries

```{r}
#library(ggplot2)
library(cowplot)
library(arm)
library(broom)
library(ggpubr)
```



## Basic regression in R

Linear regression is done in R with the lm() function, which stands for "linear model".  Most "statistical tests are or can be formulated as linear models; t-tests, ANOVA, ANCOVA, linear regression, and multiple linear regression are all linear models

Some common methods can't be fit directly as a linear model, like Welch's t-test, which corrects for unequal variance.  However, a very similar approach can be implemented using the function gls() in the nlme package.  Standard linear regression is fit using least squares methods, and "gls" stands for "generalized least squares" and makes it possible to relax many of the standard assumptions of classic statistical tests, such as constant variance and independence of residuals.

An extension of the lm() function is glm(), which stands for "general linear model" or "generalized linear model."  Linear models work with numeric data and assumes that residuals are normally distributed.  GLMs allow you to work binomial, count, and other forms of data that have non-normal residuals.  


### Data preparation

The data has been cleaned in previous tutorials.  One thing that will make modeling these data is to transform the predictor variable.  This is because the mass of animals varies so much.

If we look at the raw data its pretty ugly

```{r first.ggscagtter.chunk}
ggscatter(data = milk_primates,
         y = "fat",
         x = "mass.fem")
```

This is because the dataset contains Gorillas, which weigh almost 100,000 grams.  The next largest animals is a baboon, then a beaver (humans are absent from the dataset for some reason).

```{r}
#determine order
i. <- order(milk_primates$mass.fem,
            decreasing = T)

#display largest 3 spp
milk_primates$spp[i.][1:3]
```


Log transformation makes things more linear.

```{r}
#this could be done with dplyr::mutate() too
milk_primates$mass.fem.log <- log(milk_primates$mass.fem)
```

Now check out the plot
```{r second.ggscatter.chunk}
ggscatter(data = milk_primates,
         y = "fat",
         x = "mass.fem.log")
```


### Building a regression model

Let's fit a simple linear regression model.  This will have the form we are familiar with from geometry class of

> y = m*x + b

Where b is the y intercept and m is the slope.

In stats, we typically change the order

> y = b + m*x

Additionally, we call b "Beta 0" and m "Beta 1"

> y = B0 + B1*x

To make life hard, R will call B0 "(Intercept)".

For our particularly situation, we have the model

> fat = B0 + B1*mass.fem.log

and we want R to estimate the intercept B0 and the slope B1.

We therefore tell R

> fat ~ 1 + mass.fem.log

Where the 1 represents the intercept.  In most cases we drop it and just write

> fat ~ mass.fem.log

The model is run like this:

```{r}
lm.mass <- lm(fat ~ mass.fem.log,
             data = milk_primates)
```


## Model summary

* Look at model output with summary()
* We can also load the broom package and use the tidy() command to get a more compact output
* A function is arm::display
* we can get just the coefficients/slopes with coef() 


### Standard summary

The standard summary command gives you lots of info
```{r}
summary(lm.mass)
```

The coef() command just gives you the intercept (B0) and the slope (B1), though it doesn't label them that well.
```{r}
coef(lm.mass)
```

### broom::tidy

A cleaner output can be created with the tidy() function in the broom package:
```{r}
broom::tidy(lm.mass)
```

### arm::display

I also like arm::display()
```{r, echo = T}
arm::display(lm.mass)
```



## Plot model 

For simple linear regression with a single predictor variable we can plot things easily using the ggscatter() function from ggpubr with 'add = "reg.line"' and 'conf.int = TRUE'

```{r}
ggscatter(data = milk_primates,
          y = "fat",
          x = "mass.fem.log",
          add = "reg.line",
          conf.int = TRUE)
```


Similar functionality with more options can be had using basic ggplot.

```{r, eval = F}
ggplot(data = milk_primates,
          aes(y = fat,
              x = mass.fem.log)) +
         geom_point() +
         geom_smooth(method = lm)
```

 
The mammalsmilk package also includes a function, gg_lm_plot(), which is fairly compact and adds some annotation.  It can show either summary stats
```{r}
gg_lm_plot(lm.mass,
           xlab. = "log(female mass)",
           ylab. = "Fat %",
           annotate. = "summary")
```

Or the equation for the model
```{r, echo = T}
gg_lm_plot(lm.mass,
           xlab. = "log(female mass)",
           ylab. = "Fat %",
           annotate. = "equation")
```






```{r}

```





