
---
title: "Multiple regression with categorical variables"
author: "brouwern@gmail.com"
date: "January 30, 2017"
output:
  word_document:
    reference_docx: rmarkdown_template.docx
    fig_width: 9
    fig_height: 5
    fig_caption: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, strip.white = FALSE, cache = TRUE)
```


```{r, echo = F}
#load data
milk2 <- read.csv(file = "Skibiel_clean_milk_focal_column.csv")
milk3 <- read.csv(file = "Skibiel_clean_milk_focal_genera.csv")

library(arm)
milk2$fat.logit <- logit(milk2$fat.percent/100)

```


# Adding Categorial Variables

* The most basic regression has a continous y variable and a continuous x variable
* That is, continous response and continous predictor
* You can have multiple continous predictors, resulting in multiple regression
* You can also add categorical predictors
* Also known as factors
* Fitting a regression w/ categorical predictors is kind of like fitting 2 seperate regression models, one to each subset of data
* (However, fitting a regression w/ 2+ continous predictors is NOT like fitting 2 sep. regression1)

**Milk data's dategorical/factor/grouping variables**

* diet     
* arid
* biome


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

# Subsetting data in R

* many ways to subset data in R
* subset() is handy
* (can also use which() command)

Two biomes:
```{r}
summary(milk2$biome)
```

<br>

```{r}
#note use of "biome ==" w/ 2 "="
aquatics <- subset(milk2, 
                   select = c("fat.logit",
                              "biome",
                              "lacat.mo.NUM"),
                   biome == "aquatic")

terrest <- subset(milk2, 
                   select = c("fat.logit",
                              "biome",
                              "lacat.mo.NUM"),
                   biome == "terrestrial")
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Look at data subsets

Aquatic species
```{r}
dim(aquatics)
summary(aquatics)
```

Terrestrial species
```{r}
dim(terrest)
summary(terrest)
```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Fit seperate model to each subgroup

Model aquatic
```{r}
lm.aquatics <- lm(fat.logit ~ log(lacat.mo.NUM), 
              data = aquatics)

summary(lm.aquatics)$coefficients
```

Model terrestrials
```{r}
lm.terrest <- lm(fat.logit ~ log(lacat.mo.NUM), 
              data = terrest)

summary(lm.terrest)$coefficients
```


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Compare Model output

* 1^st^ line shows intercepts from both models
* 2^nd^ line shows slopes
* Different modes in columns
* What's different

```{r, echo = F}
library(car)
library(pander)
temp <- compareCoefs(lm.aquatics, lm.terrest)
attributes(temp)$dimnames[[2]][1] <- "Aqua est"
attributes(temp)$dimnames[[2]][2] <- "Aqua SE"
attributes(temp)$dimnames[[2]][3] <- "Terr est"
attributes(temp)$dimnames[[2]][4] <- "Terr SE"

pander(temp)
```



<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

## Plot our 2 seperate models

```{r, echo  = F}
source("./scripts_Lecture4/sx_mtext_plot_coefs_in_margin.R")

par(mfrow = c(1,2),
    mar = c(3,3,4,0))

xlim. <- range(log(milk2$lacat.mo.NUM)) 
ylim. <- range(milk2$fat.logit, na.rm = T)

#Plot aquatics
plot(fat.logit ~ log(lacat.mo.NUM),
     data = aquatics,
     #lab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     xlim = xlim.,
     ylim = ylim.,
     main = "")
abline(lm.aquatics, col = 3, lwd = 3)
m.coefs(lm.aquatics, plot.p = F)
mtext(text = "Log(lactation dur)",
      line = 2,
      side = 1)
mtext(text = "logit(percent fat)",
      line = 2,
      side = 2)
mtext(text = "Aquatic species",
      side = 1, line = -1)

#Plot terrestrials
plot(fat.logit ~ log(lacat.mo.NUM),
     data = terrest,
     #lab = "Log(lactation dur.)",
     pch = 18,
     xlim = xlim.,
     ylim = ylim.,
     cex =2,
     main = "")
abline(lm.terrest, col = 4, lwd = 3)
m.coefs(lm.terrest, plot.p = F)
mtext(text = "Log(lactation dur)",
      line = 2,
      side = 1)
mtext(text = "Terrestrial species",
      side = 1, line = -1)



```



<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

## Plot our 2 seperate models
### Compare intercepts

```{r, echo  = F}
source("./scripts_Lecture4/sx_mtext_plot_coefs_in_margin.R")

par(mfrow = c(1,2),
    mar = c(3,3,4,0))

xlim. <- range(log(milk2$lacat.mo.NUM))
ylim. <- range(milk2$fat.logit, na.rm = T)

#Plot aquatics
plot(fat.logit ~ log(lacat.mo.NUM),
     data = aquatics,
     #lab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     xlim = xlim.,
     ylim = ylim.,
     main = "")
abline(lm.aquatics, col = 3, lwd = 3)
m.coefs(lm.aquatics, plot.p = F)
mtext(text = "Log(lactation dur)",
      line = 2,
      side = 1)
mtext(text = "logit(percent fat)",
      line = 2,
      side = 2)
mtext(text = "Aquatic species",
      side = 1, line = -1)
abline(v = 0, col = 2)
arrows(x0 = 0,x1 = -1.45,
       y0=-0.1755,y1 = -0.1755,
       col = 2,angle = 10)

#Plot terrestrials
plot(fat.logit ~ log(lacat.mo.NUM),
     data = terrest,
     #lab = "Log(lactation dur.)",
     pch = 18,
     xlim = xlim.,
     ylim = ylim.,
     cex =2,
     main = "")
abline(lm.terrest, col = 4, lwd = 3)
m.coefs(lm.terrest, plot.p = F)
mtext(text = "Log(lactation dur)",
      line = 2,
      side = 1)
mtext(text = "Terrestrial species",
      side = 1, line = -1)
abline(v = 0, col = 2)
arrows(x0 = 0,x1 = -1.45,
       y0=-2.055,y1 = -2.055,
       col = 2, angle = 10)


```



<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Run single model with categorical predictors

* Adding just a categorical predictor adds a new term
* The model defines 2 lines with 3 terms

```{r}
milk2$log.lact.mo <- log(milk2$lacat.mo.NUM)
lm.lac.dur.add.biome <- lm(fat.logit ~ log.lact.mo+ 
                             biome, 
                           data = milk2)

summary(lm.lac.dur.add.biome)
```


**Model 1:**
fat.logit ~ (Int.) +                    slope*log(lactation)

**Model 2:**
fat.logit ~ (Int.) + biometerrestrial + slope*log(lactation)


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

## Translating equations

* Adding just a categorical predictor adds a new term
* The model defines 2 lines with 3 terms

**Word equations**
**Model 1:**
fat.logit ~ (Int.) +                    slope*log(lactation)

**Model 2:**
fat.logit ~ (Int.) + biometerrestrial + slope*log(lactation)

**Numeric equations**
**Model 1:**
fat.logit ~ 0.0157 +                    -0.41*log(lactation)

**Model 2:**
fat.logit ~ 0.0157 +            -2.10 + -0.41*log(lactation)


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Plot both lines together

```{r, echo = F}
source("./scripts_Lecture4/fnxn_plot_intxns.R")
```


```{r, echo = F}
par(mfrow = c(1,1),
    mar = c(4,4,2,1))
plot.interactions(model = lm.lac.dur.add.biome,
                  x.cont = "log.lact.mo",
                  x.cat = "biome",
                  y = "fat.logit",
                  cols = 2:3)
mtext("lm(fat.logit ~ lacat.mo.log + biome)", cex = 1.4)
legend("bottomleft",col = 2:3,legend = c("Aqua", "Terr"),
       lty = 1)


```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## What determines the distance betwen the two lines?

```{r}
library(arm)
display(lm.lac.dur.add.biome)
```

```{r, echo = F}
par(mfrow = c(1,1),
    mar = c(4,4,1,1))
plot.interactions(model = lm.lac.dur.add.biome,
                  x.cont = "log.lact.mo",
                  x.cat = "biome",
                  y = "fat.logit",
                  cols = 2:3)
#mtext("lm(fat.logit ~ lacat.mo.log + biome)", cex = 1.4)
legend("bottomleft",col = 2:3,legend = c("Aqua", "Terr"),
       lty = 1)
int <- coef(lm.lac.dur.add.biome)[1]
offset.  <- coef(lm.lac.dur.add.biome)[3]
temp <- int+-offset.+offset./2
arrows(x0 =0, x1 = 0,code = 3,angle = 10,lwd = 3,
       y1 = int,y0 =int+offset.)
text(labels = round(offset.,3),x = -0.5,y = -temp,cex = 1.25
     )
#m.coefs(lm.lac.dur.add.biome, plot.p = F)

```
<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>



## Interactions:

* An interaction is when one variable changes the values of another
* For a continous x categorical interaction, the slope of the continuous variable depends on the value of the categorical var.



## Model lactation duration * biome interaction

```{r}
lm.lac.dur.mult.biome <- lm(fat.logit ~ log(lacat.mo.NUM) +
                              biome +
                              log(lacat.mo.NUM)*biome, 
                            data = milk2)

summary(lm.lac.dur.mult.biome)

```


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

# Interaction in R

* R can figure out all the terms
* It "expands" the equation
* In equation below, I've just included "lacat.mo.log*biome"
    + no "+ biome" as previously
```{r}
lm.lac.dur.mult.biome <- lm(fat.logit ~ log.lact.mo*biome, 
                            data = milk2)

summary(lm.lac.dur.mult.biome)

```




<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

## Plot interactions

* Both slopes change
* Both intercepts change

```{r}

plot.interactions(model = lm.lac.dur.mult.biome,
                  x.cont = "log.lact.mo",
                  x.cat = "biome",
                  y = "fat.logit",
                  cols = 2:3)


```

<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>

## Compare models
### Significance test

* Test significance of 2nd intercept 
```{r, echo = -1}
logit.lac.duration.all <- lm(fat.logit ~ log(lacat.mo.NUM), 
                             data = milk2)
anova(lm.lac.dur.add.biome, 
      logit.lac.duration.all)
```


* We can test the significance of the interaction

```{r}
anova(lm.lac.dur.mult.biome, 
      lm.lac.dur.add.biome)
```


<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## Testing all terms at once

* test with the anova() command on a single model

```{r}
anova(lm.lac.dur.mult.biome)
```



<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>


## AIC

* AIC goes down by 318.0-317.7=`r 318.0-317.7`
* This is a very very slight improve

```{r,echo = -c(1:2)}
library(bbmle)
logit.null.alldat <- lm(fat.logit ~ 1, data = milk2)

AICtab(logit.null.alldat,
       lm.lac.dur.add.biome, 
       lm.lac.dur.mult.biome,
      logit.lac.duration.all,base = T)
```




<br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br><br>



# Bayesian
```{r}

# library(MCMCglmm)
# 
# 
# lm.lac.dur.MCMC <- MCMCglmm(fat.logit ~ log(lacat.mo.NUM),
#                             data = milk2[,])
# 
# plot(lm.lac.dur.MCMC)
```



```{r}
# display(digits = 8)
# milk3 <- milk3
# lm.lac.dur.logit <- lm(fat.logit ~ log(lacat.mo.NUM),
#                        data = milk3)
# 
# milk3$resids <- resid(lm.lac.dur.logit)
# 
# milk3$resids.sq <- milk3$resids^2
# 
# sum(milk3$resids.sq)


```

 
