
---
title: "Regression: model diagnostics in action"
author: "Nathan Brouwer | brouwern@gmail.com | @lowbrowR"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, strip.white = FALSE, cache = TRUE)
```


```{r, echo = F}
#load data
milk2 <- read.csv(file = "Skibiel_clean_milk_focal_column.csv")
milk3 <- read.csv(file = "Skibiel_clean_milk_focal_genera.csv")
```



## All milk fat data


```{r, echo = F}
par(mfrow = c(1,2), 
    mar = c(4,4,2,1))

#standard plot
plot(fat ~ log(lac.mo),
     data = milk,
     #lab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     main = "plot()")


#scattersmooth()

with(milk2, scatter.smooth(y= fat.percent, 
     x = log(lacat.mo.NUM),
     main = "scatter.smooth()",
     lpars = list(col = 3, lwd = 3)))
```

* There's an important feature of these data that I haven't mentioned ...
* It has major statistical implications
* Relates to the variable fat.percent



## An important feature of these data


```{r, echo = F}
par(mfrow = c(1,2), mar = c(4,4,2,1))
ylim. <- c(0,100)
#standard plot
plot(fat.percent ~ log(lacat.mo.NUM),
     data = milk2,
     #lab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     ylim = ylim.,
     main = "plot()")


#scattersmooth()

with(milk2, scatter.smooth(y= fat.percent, 
     x = log(lacat.mo.NUM),
     main = "scatter.smooth()",
          ylim = ylim.,
     lpars = list(col = 3, lwd = 3)))
```

* There's an important feature of these data that I haven't mentioned ...
* It has major statistical implications
* Relates to the variable fat.percent






## An important feature of these data: 
### Percentages are bounded between 0% and 100%


```{r, echo = F}
par(mfrow = c(1,2), mar = c(4,4,2,1))
ylim. <- c(0,100)
#standard plot
plot(fat.percent ~ log(lacat.mo.NUM),
     data = milk2,
     #lab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     ylim = ylim.,
     main = "plot()")
abline(h = 1,lwd = 4, col = 2)
abline(h = 100,lwd = 4, col = 2)


#scattersmooth()

with(milk2, scatter.smooth(y= fat.percent, 
     x = log(lacat.mo.NUM),
     main = "scatter.smooth()",
          ylim = ylim.,
     lpars = list(col = 3, lwd = 3)))
abline(h = 1,lwd = 4, col = 2)
abline(h = 100,lwd = 4, col = 2)

```






## Percentages are bounded between 0% and 100%

* A regression model won't respect this
* We can try to account for this using "quadratic term" 
  + square the predictor variable
  + fat ~ log(lacat.mo.NUM) + log(lacat.mo.NUM)^2
  


### Fit quadratic model

* Syntax for adding a squared term is annoying
    + Wrap squared term in "I(...)"
    + lm(fat.percent ~ log(lacat.mo.NUM) + I(log(lacat.mo.NUM)^2)
* Not: squaring a term that is already logged results in a somewhat strange model when you think about the math...

```{r}
#Null model
lm.null.alldat <- lm(fat.percent ~ 1, data = milk2)


#refit lactation duration model to all data
lm.lac.dur.alldat <- lm(fat.percent ~ log(lacat.mo.NUM), data = milk2)

#add squared term
lm.lac.dur.squared.alldat <- lm(fat.percent ~    
                        log(lacat.mo.NUM) +                        I(log(lacat.mo.NUM)^2), 
                                data = milk2)

```



# Look at diagnostics of linear model

```{r, echo =-1}
par(mfrow = c(2,2))
plot(lm.lac.dur.alldat)
```




# QQ plot for normality is horrible!

```{r, echo =-1}
par(mfrow = c(1,1))
plot(lm.lac.dur.alldat, which = 2)
```





# Does including x^2^ quadratic term improve model?

* Maybe a tiny bit...

```{r, echo =-1}
par(mfrow = c(1,1))
plot(lm.lac.dur.squared.alldat, which = 2)
```
`




### Summary of model with square term

* Note "I(log(lacat.mo.NUM)^2)" term
* This is highly significant

```{r}
summary(lm.lac.dur.squared.alldat)
```



### Compare models w/ anova()

* The original and squared model are "nested"
* Test with anova()
* x^2 term is significant

```{r}
anova(lm.lac.dur.squared.alldat,
      lm.lac.dur.alldat)
```






### Compare models w/ AIC

* Can compare multiple models easily w/AIC
    + Null: . ~ 1
    + Alt1: . ~ log(lac.dur)
    + Alt2: . ~ log(lac.dur) +log(lac.dur)^2

```{r, echo = -1}
library(bbmle)
AICtab(lm.lac.dur.squared.alldat,
      lm.lac.dur.alldat,
      lm.null.alldat)
```





### Look at model fit 

* Does the model fit the data well?

```{r, echo = F}
#plot
par(mfrow = c(1,2), mar = c(4,4,2,1))
ylim. <- c(0,100)

# plot linear model
plot(fat.percent ~ log(lacat.mo.NUM),
     data = milk2,
     #lab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     ylim = ylim.)
mtext(text = "Linear: fat ~ log(lac)")
abline(h = 1,lwd = 4, col = 2)
abline(h = 100,lwd = 4, col = 2)
abline(lm.lac.dur.alldat, col = 3, lwd = 3)

# plot quadratic model
plot(fat.percent ~ log(lacat.mo.NUM),
     data = milk2,
     #lab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     ylim = ylim.)
mtext(text = "Non-lin: fat ~ log(lac)+log(lac)^2")
abline(h = 1,lwd = 4, col = 2)
abline(h = 100,lwd = 4, col = 2)

## plot predictions from x^2 model
preds <- predict(lm.lac.dur.squared.alldat)
col.name <- "log(lacat.mo.NUM)"
x <- lm.lac.dur.squared.alldat$model[,col.name]


#put in order so it plots right
points(preds[order(x)] ~ x[order(x)], 
       type = "l",
       col= 3,
       lwd = 3)

```




### Add cubic term

* . ~ x + x^2 + x^3
* This is generally a bad idea...
  + Results in "overfitting"
  + Hard to justify what biological process results in a x^3 term

```{r, echo = F}
par(mfrow = c(1,2))
# Add cubic term
lm.lac.dur.cubed.alldat <- lm(fat.percent ~    
                        log(lacat.mo.NUM) +                        I(log(lacat.mo.NUM)^2) +
                          I(log(lacat.mo.NUM)^3), 
                                data = milk2)

# plot quadratic model
plot(fat.percent ~ log(lacat.mo.NUM),
     data = milk2,
     #lab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     ylim = ylim.)
mtext(text = "Non-lin: fat ~ log(lac)+log(lac)^2")
abline(h = 1,lwd = 4, col = 2)
abline(h = 100,lwd = 4, col = 2)

## plot predictions from x^2 model
preds <- predict(lm.lac.dur.squared.alldat)
col.name <- "log(lacat.mo.NUM)"
x <- lm.lac.dur.squared.alldat$model[,col.name]


#put in order so it plots right
points(preds[order(x)] ~ x[order(x)], 
       type = "l",
       col= 3,
       lwd = 3)


# plot cubic model
plot(fat.percent ~ log(lacat.mo.NUM),
     data = milk2,
     #lab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     ylim = ylim.)
mtext(text = "Non-lin: fat ~ ...log(lac)^3")
abline(h = 1,lwd = 4, col = 2)
abline(h = 100,lwd = 4, col = 2)

## plot predictions from x^2 model
preds <- predict(lm.lac.dur.cubed.alldat)
col.name <- "log(lacat.mo.NUM)"
x <- lm.lac.dur.squared.alldat$model[,col.name]


#put in order so it plots right
points(preds[order(x)] ~ x[order(x)], 
       type = "l",
       col= 3,
       lwd = 3)


```




# Stop the maddness: transform percentages

* log transformations are very common for continuous data
* for percentage data, the "logit transformation" is what's needed
  + everyone we did previously should've been done w/logit transform
* people often do an "arcsine" transformation but this is being outmoded





## Implement logit transformation

* logit transformation functions are in car and arm packages
* transform bounded percentages [0,100] to unbounded variables
* the math is in the same next of the woods as logistic regression

```{r}
library(arm)
milk2$fat.logit <- logit(milk2$fat.percent/100)

```




## Compare distribution of the variables

* Improves normality of raw data
* (note: we really care about normality of residuals!)

```{r, echo = -1}
par(mfrow = c(1,2))
hist(milk2$fat.percent,main = "fat.percent")
hist(milk2$fat.logit, main = "logit(fat.percent)")
```





### Model logit and compare to raw percent

* We can't compare logit and raw percentages w/ a significance test b/c the models aren't nested
* can use AIC

```{r, echo = F}
# Model
logit.lac.duration.all <- lm(fat.logit ~ log(lacat.mo.NUM), 
                             data = milk2)
```

```{r, echo = F}
summary(logit.lac.duration.all)

```





```{r, echo = F}
par(mfrow = c(1,1))
plot(fat.logit ~ log(lacat.mo.NUM),
     data = milk2,
     #lab = "Log(lactation dur.)",
     pch = 18,
     cex =2,
     main = "logit(fat.percent) ~ log(lactation duration)")
abline(logit.lac.duration.all, col = 3, lwd = 3)

```



# Diagnostics for logit model

```{r}
par(mfrow = c(2,2))
plot(logit.lac.duration.all)
```



# qqplot Diagnostics for logit model

```{r}
par(mfrow = c(1,1))
plot(logit.lac.duration.all, which = 2)
```



