---
title: "Duq Biostats Week 5 | Regression Tutorial 6: Model selection"
author: "Nathan Brouwer | brouwern@gmail.com | @lowbrowR"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, strip.white = FALSE, warning = FALSE,message = FALSE)
```



```{r, echo = FALSE}
setwd("C:/Users/lisanjie2/Desktop/TEACHING/1_STATS_CalU/1_STAT_CalU_2016_by_NLB/1_STAT_Duq_2017/HOMEWORK/Duq_Biostats_Homework_4_Milk_regression/homework4_RMD_vs2")
dat2 <- read.csv("milk_subset.csv")
```


# Introduction

## Original Data

**Skibiel et al 2013.**  Journal of Animal Ecology.  The evolution of the nutrient composition of mammalian milks. 82: 1254-1264.


## Model selection tutorial: Relative lactation length

* We saw that body size is not an important variable for predicting milk composition
* a variable that is important in most of the models Skibiel et al 2013 fit is "relative lactation duration." or "relative lactation length" 



They define relative lactation duration as:

**"Lactation length was analysed relative to the total period of reproduction so that relative lactation length = absolute lactation length/(absolute gestation length + absolute lactation length)."**

* This transformation to a proportion corrects for the fact that 
    + longer gestation results in larger offspring 
    + and hence a longer periods of lactation as these large babies mature.
* We'll look in depth at their models for % fat~lactation duration to understand model selection and model comparison.




# Data setup

## Relative lactation duration

Create a variable for relative lactation length.  Use the **with()** command to make the code clean

```{r}
dat2$rel.lactation <- with(dat2, lacatation.months/(gestation.month +                                             lacatation.months))
```



## Transformation

* As the authors did, Log10 transform relative lactation
* (recall that log = natural log in R)

```{r}
dat2$rel.lact.log10 <- log10(dat2$rel.lactation)
```


Relative Lactation duration is a proportion, so a logit transformation may be more appropriate.

```{r}
#the car package has a logit function
library(car)

#apply logit transformation
dat2$rel.lact.logit <- logit(dat2$rel.lactation,percents = T)
```




# Data exploration

## Untransformaed dat: Plot log10(fat) ~ rel.lactation

* First, plot data w/ No transformation of the predictor variable.
* Note the triangle shape of the data.  This forebodes heterogeneous variance.

```{r}
library(ggplot2)

qplot(y = fat.log10,
      x = rel.lactation,
      data = dat2) + 
  ggtitle("no transformation of x")
```


## Author's transformation: Plot log10(fat) ~ log10(rel.lactation)

* Interesting.  This makes the plot really ugly.
* An un-transformed x axis might be better


```{r}
qplot(y = fat.log10,
      x = rel.lact.log10,
      data = dat2)+ 
  ggtitle("log10 transformation of x")
```


## Plot log10(fat) ~ logit(rel.lactation)

* "rel.lactation" is a proportion so perhaps the logit would be better
* Logit makes it look bad too!

```{r}
qplot(y = fat.log10,
      x = rel.lact.logit,
      data = dat2)
```



## Plot logit(fat) ~ logit(rel.lactation)

* logit(y)~logit(x)
* Still ugly.  
* Any transformation of the x variable seems bad.  
* I believe the original authors did do use the log transform of x axis, but perhaps they didn't check what it did and just assumed it was a good idea.
    + I can't totally reproduce their results so I'm not 110% sure that they did.

```{r}
qplot(y = fat.logit,
      x = rel.lact.logit,
      data = dat2)
```



## Look at different transformations

Fit models of un-transformed lactation duration and compare to log and logit transformations

```{r}
#model 1: log
f.rellact.raw <- lm(fat.log10 ~ rel.lactation, data = dat2)


#modlel 2: log-log
f.rellact.log10 <- lm(fat.log10 ~ rel.lact.log10, data = dat2)


#model 3: log-logit
f.rellact.logit <- lm(fat.log10 ~ rel.lact.logit, data = dat2)

```


The anova() command gives us the p value for the slope
```{r}
anova(f.rellact.raw)
anova(f.rellact.log10)
anova(f.rellact.logit)
```

We can't compare these 3 models directly, but we can assess how well the fit the data by looking at their residuals


### Look at residuals

#### Residuals for the raw data:

Resid vs. fitted looks pretty bad - definite triangle shape.
```{r}
par(mfrow = c(2,2))
plot(f.rellact.raw)
```




#### Residuals for the log and logit

Resid vs. fitted looks pretty bad - can't say its really worse necessarily, but my intuition is that it is.  

```{r}
plot(f.rellact.log10)
```

```{r}
plot(f.rellact.logit)
```


* These diagnostic plots indicate that there may be problems with heterogeneous variance
* the problem may be worse using log transformed relative lactation duration as a predictor (that is, a long transformed x axis).
* However, we are not using all of the predictors yet that the the author's did. 
* Adding these additional predictors _may_ alleviate the issues in the data.
* In general you want to do diagnostics on your best model to confirm that it meets the assumptions
* However, exploring the residuals of all your models can help you better understand the data




# Model selection

* We'll now fit a series of models as the author did 
* then use AIC to assess which one is best.  
* They used relative lactation duration as a continuous predictor of the fat content of milk, & also several categorical variables: 
    + biome (terrestrial vs. aquatic)
    + diet (carnivore, herbivore, omnivore)
    + arid habitat (yes/no)  

They looked at different combinations of these predictors


## Fit models

Fit models with all of the combinations of these predictors - this is a long list


### Null model
```{r}
#Model w/no predictors
###Note: authors don't fit this model
f.null <- lm(fat.log10 ~ 1,
             data = dat2)

```

### Models w/ 1 predictor

```{r}
#continous predictor
##rel lactation length (continous)
f.rellact <- lm(fat.log10 ~ rel.lact.log10,
                data = dat2)

#categorical predictor
##Biome (categorical)
f.biome <- lm(fat.log10 ~ biome,
              data = dat2)

##Diet (cat.)
f.diet <- lm(fat.log10 ~ diet, 
             data = dat2)

##Arid-adapted (cat)
f.arid <- lm(fat.log10 ~ arid,
             data = dat2)


```



### Models w/2 predictors

* I often call these "additive" models because they imply a hypothesis that the two predictors have seperate effects
* Also because they are defined with a "+" sign
  + This is in contrast to an interaction model, defined by multiplication ("*")

```{r}

#Biome + diet (2 categorical vars)
f.biome.diet <- lm(fat.log10 ~ biome + 
                               diet,
                   data = dat2)

#Rel lac length + diet (1 continous, 1 cat)
f.lact.diet <- lm(fat.log10 ~ rel.lact.log10 + 
                    diet,
                  data = dat2)


#Rel lac length + biome (1 cont, 1 cat)
f.lact.biome <- lm(fat.log10 ~ rel.lact.log10  +
                    biome, 
                  data = dat2)

```


### Model w/3 terms

* Continuous: rel.lact.log10
* Categorical: biome, and diet

```{r}
#Re lac  + diet + biome (1 cont, 2 cat)
f.lact.biome.diet <- lm(fat.log10 ~    
                          rel.lact.log10 + 
                          biome +
                          diet, 
                          data = dat2)
```



# Model selection w/AIC


## Get a single AIC value

We can get individual AIC values for a model using AIC()

```{r}
AIC(f.biome)
```

**QUERSTION:**
How do you interpret an AIC value?























**ANSWER:**
You can't; on its own, and AIC value is meaningless


## Get an AIC table

* We can get ALL AIC values using bbmle:AICtab.  
* "base=TRUE" returns the raw AIC value in addition to the delta AIC. 

```{r}
library(bbmle)
AICtab(base= TRUE,
       f.null,
       f.biome,
       f.rellact,
       f.diet,
       f.arid,
       f.biome.diet,
       f.lact.diet,
       f.lact.biome,
       f.lact.biome.diet
)
```



## Get an AICc table

* To get AICc values (small sample correction) we use the ICtab function
* Its annoying that there isn't AICctab()
* They guy that wrote bbmle, Ben Bolker, is so awesome I'll let it slide


```{r}
ICtab(type = "AICc",
      base = TRUE,
      f.null,
       f.biome,
       f.rellact,
       f.diet,
       f.arid,
       f.biome.diet,
       f.lact.diet,
       f.lact.biome,
       f.lact.biome.diet)
```


* These AIC values are similar in magnitude as the authors report, but not identical.  
* (F statistics they report are also similar to what I get, but not identical).  
* I am not sure what the issue is.  I think their best model had an AICc of 75ish, while I get 97.8.


## Check the impact of different transformations

* Our best model, "f.lact.biome.diet", was fit like this 
    + both response and predictor variables are on log scale

```{r}
#Re lac  + diet + biome (1 cont, 2 cat)
f.lact.biome.diet <- lm(fat.log10 ~    
                          rel.lact.log10 + 
                          biome +
                          diet, 
                          data = dat2)
```


Let's fit a model with %fat transformed on the logit scale

```{r}
f.lact.biome.diet.LOGIT <- lm(fat.logit ~    
                          rel.lact.log10 + 
                          biome +
                          diet, 
                          data = dat2)
```


### COmpare different transformation w/AIC

```{r}
AICtab(base = TRUE,
       f.lact.biome.diet,
       f.lact.biome.diet.LOGIT)
```

* AIC for the log(fat) model is MUCH MUCH lower that logit(fat) model we just fit
* Does this mean that log(fat) is  MUCH MUCH better?
* NO!  It means that we fit models to two datasets on very very different scales?



**QUESTION**

* What might be the impact of having missing data (As) for some of your predictors but not others?  
    + For example, what if you had "biome" data for every mammals species
    + but you don't have diet info for 10 of the species
    + When you lack data for a predictor and include that predictor in the model, R drops the entire row of data from the analysis
    






















**ANSWER**

* When you have missing data for some predictors you cannot compare models w/AIC
* anova() will usually give you a warning; AIC() and AICtab won't!
* You can only compare models that are fit to the exact same dataset
* I suspect this is a major problem that people forget about.


# Exploring the impact of missing data

Let's make a new dataset from our milk  data and create some missing data

```{r}
dat.fake <- dat2
```


I'll randomly remove 5 observations from just the "diet column"
```{r}
#runif() draws random numbers from a uniform distribution
i.random <- runif(n = 5, min = 1, max = length(dat2$diet))

i.random <- round(i.random)

#Change random variables to NA
dat.fake[i.random, "diet"] <- NA

```


Now I"ll fit a model with rel.lact.log10 and diet to the original data
```{r}
#just relative lacation
m.rel.lac <- lm(fat.logit ~ rel.lact.log10, 
                          data = dat2)

#relative lacation and diet
m.rel.lac.diet <- lm(fat.logit ~ rel.lact.log10 +
                       diet, 
                          data = dat2)

```


Now I'll fit a model to the faked data

```{r}
#just relative lacation
fake.m.rel.lac <- lm(fat.logit ~ rel.lact.log10, 
                          data = dat.fake)

#relative lacation and diet
fake.m.rel.lac.diet <- lm(fat.logit ~ rel.lact.log10 +
                            diet, 
                          data = dat.fake)

```



Compare models with correct data

```{r}
AICtab(base = TRUE,
       m.rel.lac,
       m.rel.lac.diet)
```



Compare models with faked  data

```{r}
AICtab(base = TRUE,
       fake.m.rel.lac,
       fake.m.rel.lac.diet)
```



The original and "fake" models with just lactation duration as a predictor are the same b/c they are fit to the same dataset
```{r}
AICtab(base = TRUE,
       m.rel.lac,
       fake.m.rel.lac)
```




The original and "fake" models with "diet" add have different AICs because they are fit to different data
```{r}
AICtab(base = TRUE,
       m.rel.lac.diet,
       fake.m.rel.lac.diet)
```



We will only see an error if we use AICc.  This is because AICc incorporates sample size into its equation

```{r, echo = F}
ICtab(base = TRUE,
      type = "AICc",
       m.rel.lac.diet,
       fake.m.rel.lac.diet)
```


We'll also see an error if we try anova() on the two fake models
```{r}
#test "fake"
anova(fake.m.rel.lac,
       fake.m.rel.lac.diet)
```




# Model Diagnostics:

* Let's look at diagnostic plots of  the best model, f.lact.biome.diet

```{r}
AIC(f.lact.biome.diet)

plot(f.lact.biome.diet)
```

**QUESTION**
What do you thin about these residuals?























**ANSWER**
* The resid vs. fitted model is still pretty bad;
* addition of biome and diet as predictors do not improve the model's fit.


# Plotting model output

* The best way to check a model is to plot it against the raw data. 
* To make our life simpler we'll just consider the model that had lactation duration and diet 
    + That is, we'll ignore the "biome" predictor.  
* This will take a bit of work because the categorical predictor has three levels
* In general ggplot() has better facilities for doing this, but the syntax is very different from plot()
* Also setting this up is good practice for basic R skills


## Seperating indices for each plot element

First, I'll create indices for the  3 diet groups using which()

```{r}
#indices for each diet group
i.omni <- which(dat2$diet == "omnivore")
i.carn <- which(dat2$diet == "carnivore")
i.herb <- which(dat2$diet == "herbivore")

```


## Plot different diets

* Next I'll plot all the data in black using plot(), 
* then overlay carnivores in red (col = "red" or col = 2)
* and herbivores in green (col = "green" or col = 3)
* use points() to add new elements to a plot
* Note use of indices to select the subset of data being used
    + data = dat2[i.carn,] for carnivores
    + data = dat2[i.herb,] for herbivores
* Legend gets added with length()
    + ggplot also makes pretty good legend on the fly
    + why don't I just invest the effort to each ggplot from the beginning...

```{r}
#plot all raw data
par(mfrow = c(1,1))
plot(fat.log10 ~ rel.lact.log10, 
     data = dat2)

#color carnivores
points(fat.log10 ~ rel.lact.log10, 
       data = dat2[i.carn,],
       col = "red")

#color herbivores
points(fat.log10 ~ rel.lact.log10, 
       data = dat2[i.herb,],
       col = "green")


#add a legend
legend("bottomleft", 
       legend = c("Carnivore", 
                  "Omnivore",
                  "Herbiore"),
       col = c("red",
               "green",
               "black"),
       pch = 1)

```


## Add regression lines

* To plot the regression lines I'll generate predictions (y.hat values) using predict().  
* This uses the regression equation from the specified model (f.lact.biome.diet) and plugs in observed values (xs) for the 3 predictors (relative lactation duration, biome, and diet).
* This requires a bit of code b/c there are 3 levels to the diet variable


### Generate predictions

Generate predictions for plotting works best if we create a new dataframe with organized data the span the entire original dataset.



The range of lactation duration values
```{r}
# ranges uses min() and max()
rng <- range(dat2[,"rel.lact.log10"])

```



Levels of the factor variables.  

* These are accessed with the levels() command
```{r}
#levels of the diet factor
levs1 <- levels(dat2[,"diet"])

levs1

#levels of the biome fator
levs2 <- levels(dat2[,"biome"])
  

```


Create a new dataframe from which to generate predictions

* This uses the expand.grid function
* This generates all combinations of the variables that are given

```{r}
newdat <- expand.grid(rel.lact.log10 = rng,
                      diet = levs1,
                      biome = levs2)

newdat
```



Note: I'm using the model with lactation and diet (f.lact.diet), NOT the one best model which also included "biome" (f.lact.biome.diet).

### Generate prediction

* use predict()
* give it the model in question, "f.lact.diet""
* give it the new data we made with expand.grid()
    + y.hat is the fancy name for model predictions


```{r}
y.hat <- predict(f.lact.diet,
                 newdata = newdat)
```


**POP QUIZ**
y.hat - y = ??????
























**ANSWER***
y.hat - y = residuals


Add the y.hat values (predictions) to the newdat dataframe

```{r}
newdat$fat.y.hat <- y.hat
```



## Plot everything

* Now plot the lines define by the y.hat values
* Since we're using RMarkdown we have to include both the code to plot the points in the chunk, followed by the code for the lines. 
* Note that the regression lines will be plotted with points() also, with the argument "type = 'l'" for lines.



f.null,
       f.biome,
       f.rellact,
       f.diet,
       f.arid,
       f.biome.diet,
       f.lact.diet,
       f.lact.biome,
       f.lact.biome.diet
       

```{r}
#Plot points
## plot base plot
plot(fat.log10 ~ rel.lact.log10, data = dat2)


##plot red points
points(fat.log10 ~ rel.lact.log10, data = dat2[i.carn,],col = "red")

##plot green points
points(fat.log10 ~ rel.lact.log10, data = dat2[i.herb,],col = "green")


#Plot regression lines

## Indices for each subset
i.carn.newdat <- which(newdat$diet == "carnivore")
i.omni.newdat <- which(newdat$diet == "omnivore")
i.herb.newdat <- which(newdat$diet == "herbivore")

## Plot lines
### type = "l" makes points() plot lines
points(fat.y.hat ~ rel.lact.log10, 
       data = newdat[i.carn.newdat,],
       col = "red",
       type = "l")

points(fat.y.hat ~ rel.lact.log10, 
       data = newdat[i.omni.newdat,],
       col = "black",
       type = "l")

points(fat.y.hat ~ rel.lact.log10, 
       data = newdat[i.herb.newdat,],
       col = "green",
       type = "l")
```




**QUESTION**
It looks like the herbivore and omnivore lines are almost exactly on top of each other. What does this mean?































**ANSWER**
This means the must have very similar intercept parameters.  I can check this with **coef()**

```{r}
coef(f.lact.diet)
```

Both "dietherbivore" and  "dietomnivore" have intercepts of about -0.57.



**QUESTION**
How do you write this as a regression equation?






























**GENERAL MODEL STATEMENT**
y = Intercept + rel.lact.log10 + diet

The "diet" category has 3 levels


**EQUATION WITH COEFFICIENTS**
y = 1.2 + -0.45*rel.lact.log10 
        + -0.58*(Is diet herbivore?) 
        +  -0.57*(Is diet omnivore?) 


* The intercept of 1.2 corresponds to carnivores
* -0.58 is the difference between carnivores and herbivores
* -0.57 is the difference between carnivores and omnivore


# Back to heterogeneity of variance

* Our regression line for carnivores fits those species pretty well, but the omni- and herbivore lines don't fit their data well.  
* This indicates the source of our heterogeneity of variance problem:





## Visualize coefficients

Here's a trick: I"m going to put all of our plotting code into a function


```{r}
make.plot <- function(){
  #Plot points
## plot base plot
plot(fat.log10 ~ rel.lact.log10, data = dat2)


##plot red points
points(fat.log10 ~ rel.lact.log10, data = dat2[i.carn,],col = "red")

##plot green points
points(fat.log10 ~ rel.lact.log10, data = dat2[i.herb,],col = "green")


#Plot regression lines

## Indices for each subset
i.carn.newdat <- which(newdat$diet == "carnivore")
i.omni.newdat <- which(newdat$diet == "omnivore")
i.herb.newdat <- which(newdat$diet == "herbivore")

## Plot lines
### type = "l" makes points() plot lines
points(fat.y.hat ~ rel.lact.log10, 
       data = newdat[i.carn.newdat,],
       col = "red",
       type = "l")

points(fat.y.hat ~ rel.lact.log10, 
       data = newdat[i.omni.newdat,],
       col = "black",
       type = "l")

points(fat.y.hat ~ rel.lact.log10, 
       data = newdat[i.herb.newdat,],
       col = "green",
       type = "l")
}

```



Now make the plot

```{r}
make.plot()
```




Make the plot and add intercept

```{r}
#make plot
make.plot()

#add verticle line at intercept
abline(v = 0)

#get intercept term from model
int <- coef(f.lact.diet)[1]

#plot arow for intercept
arrows(y0 = int,
       x0 = -0.2,
       y1 = int,x1 = 0, 
       length = 0.1,col = 2,lwd = 3)

```



Make the plot and add intercept+"dietherbivore" term

```{r}
#make plot
make.plot()

#add verticle line at intercept
abline(v = 0)

#get intercept term from model
int <- coef(f.lact.diet)[1]
dietherbivore <- coef(f.lact.diet)[3]


#add the intercept and "dietherbivore" terms together
int.plus.dietherbivore <- int+dietherbivore

#plot arow for intercept
arrows(y0 = int,
       x0 = -0.2,
       y1 = int,x1 = 0, 
       length = 0.1,col = 2,lwd = 3)

#plot arow for intercept+dietherbivore
arrows(y0 = int.plus.dietherbivore,
       x0 = -0.2,
       y1 = int.plus.dietherbivore,
       x1 = 0, 
       length = 0.1,col = 3,
       lwd = 3)

```




# Further model and data explorations

## What about outliers identified previously

Index values of weird groups
```{r}
# 
i.aust <- which(dat2$Australia == "Australia")
i.lemur <- which(dat2$lemurs == "lemur")
i.odd.ung <- which(dat2$order == "Perrissodactyla")

```



### Plot Australia
```{r}
#make plot
make.plot()

#plot points
## australia spp
points(fat.log10 ~ rel.lact.log10, 
       data = dat2[i.aust,],
       col = "black",
       pch = 16)
```


### Plot Lemurs
```{r}
#make plot
make.plot()

##lemurs
points(fat.log10 ~ rel.lact.log10, 
       data = dat2[i.lemur,],
       col = "black",
       pch = 22)
```



### Plot odd-toed ungulates
```{r}
#make plot
make.plot()

## odd-toed ungulates
points(fat.log10 ~ rel.lact.log10, 
       data = dat2[i.odd.ung,],
       col = "black",
       pch = 3)
```



### Plot everything

* add legend with legend()

```{r}


#make plot
make.plot()


## australia spp
points(fat.log10 ~ rel.lact.log10, 
       data = dat2[i.aust,],
       col = "black",
       pch = 16)

##lemurs
points(fat.log10 ~ rel.lact.log10, 
       data = dat2[i.lemur,],
       col = "black",
       pch = 22)

## odd-toed ungulates
points(fat.log10 ~ rel.lact.log10, 
       data = dat2[i.odd.ung,],
       col = "black",
       pch = 3)




#plot legend
legend("bottomleft",pch = c(16,22,3,1), legend = c("Australia","lemur","even-toed ungulates","other"))



```

