---
title: "m) Regression Introduction"
author: "Nathan Brouwer | brouwern@gmail.com | @lowbrowR"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, strip.white = FALSE, cache = TRUE)
```


```{r, echo = F}
#load data
milk2 <- read.csv(file = "Skibiel_clean_milk_focal_column.csv")
milk3 <- read.csv(file = "Skibiel_clean_milk_focal_genera.csv")
```


```{r, echo=FALSE}
par(mfrow = c(1,1))
source("./scripts_Lecture4/sx_reg_flowchart1_inf_predict_model.R")

lm.null <- lm(fat.percent ~ 1,
              data = milk3)

```

# What about our other predictors?

* female size was just one of the predictors
    + litter size, duration of lactation, etc
* We can use hypothesis tests to determine if each one of these is "significant"
* But, can't compare these models because they aren't "nested"
* In contrats, AIC can compare "non-neste models"






## Fit models for each predictor

* All of these models have 2 parameters (slope, intercept)
* But, are NOT nested b/c the slope is for a difference predictor
* nested = "a special case of", "reduced version"
    + the null model (...~1) is nested in all of them b/c it lacks this slope parameter

```{r}
# re-fit female mass model
lm.mass <- lm(fat.percent ~ log(mass.female), data = milk3)

#gestation duration
lm.gestation <- lm(fat.percent ~ log(gest.month.NUM), data = milk3)

#lactation duration
lm.lac.duration <- lm(fat.percent ~ log(lacat.mo.NUM), data = milk3)

#Litter mass
lm.litter.mass <- lm(fat.percent ~ log(mass.litter), data = milk3)

```





## Which predictor fits the data best?

### NSHT: do significance test for each predictor against the null

```{r}
anova(lm.mass, lm.null)
anova(lm.gestation, lm.null)
anova(lm.lac.duration, lm.null)
anova(lm.litter.mass, lm.null)
```

* p lm.mass         = 0.00044 ***
* p lm.gestation    = 0.0011 **
* p lm.lac.duration = 7.5e-06 ***
* p lm.litter.mass  = 0.013 *
    + How to judge which one is best?
    + Each one is significant...
    + Lower p-value does not! mean is a better fit




### Information theory-AIC (IT-AIC)

* AIC can compare non-nested models
```{r, echo = -1}
library(bbmle)
ICtab(lm.mass,
      lm.gestation,
      lm.lac.duration,
      lm.litter.mass)
```


# What about R2?

```{r, echo = F}

AIC <- c(AIC(lm.lac.duration),
AIC(lm.mass),
AIC(lm.gestation),
AIC(lm.litter.mass))

R2  <- c(summary(lm.lac.duration)$r.squared,
summary(lm.mass)$r.squared,
summary(lm.gestation)$r.squared,
summary(lm.litter.mass)$r.squared)

slopes <- c(coef(lm.lac.duration)[2],
coef(lm.mass)[2],
coef(lm.gestation)[2],
coef(lm.litter.mass)[2])

mods <- c("Lactation duration",
          "Female mass",
          "Gestation duration",
          "Litter mass")
AIC.vs.R2.vs.slope <- data.frame(AIC,R2,slopes,
                                 row.names = mods)

```

```{r, echo = F}
library(pander)
pander(AIC.vs.R2.vs.slope)
```

* AIC correlates with R2 in this case
* Slopes not related to AIC/R2
* Its been shown that in general R2 SHOULD NOT be used for model selection/model comparison
* R2 ALWAYS goes up when parameters added to models
* Adjusted-R2 adjusted for this
* Adjusted-R2 still doesn't work, for model selection
* R2 is a descriptive statistics, NOT an inferential stat






# Can we include multiple predictors?

* R2 is low
* Lots of scatter around line, lots of variation still in data
* We can build models with more than one of these predictors
    + can do 2, 3 or more
    + can do "interactions" between predictors
* This is "multiple regression" b/c there are multiple predictors
* Each predictor gets its own slope
* (but only one intercept)


## Model with 2 predictors
```{r}
# Model model according to AIC
lm.lac.duration <- lm(fat.percent ~ log(lacat.mo.NUM), data = milk3)


#Add additional predictor
lm.lac.dur.fem.mass <- lm(fat.percent ~ log(lacat.mo.NUM) +
                            log(mass.female), 
                          data = milk3)

```




## Summary of model w/2 predictors

```{r}
summary(lm.lac.dur.fem.mass)
```

* female mass NOT significant...
* but it was on its own when run previously
* Why would this happen?




# Should we include multiple predictors?

```{r, echo = F}
par(mfrow = c(2,2))

#Mass vs lactation duration
plot(log(lacat.mo.NUM) ~ log(mass.female),
     data = milk3,
     ylab = "Log(lactation dur.",
     pch = 18,
     cex =2,
     main = "Mass vs lactation duration")
cor.x <- round(with(milk3,cor(log(lacat.mo.NUM), log(mass.female))),3)
cor. <-expression(paste(rho,"=0.75"))
mtext(text = cor.)

#Mass vs Gestation duration
plot(log(gest.month.NUM) ~ log(mass.female),
     data = milk3,
     ylab = "Log(gestation dur.)",
     pch = 18,
     cex =2,
     main = "Mass vs Gestation duration")

#maternal Mass vs Litter mass
plot(log(mass.litter) ~ log(mass.female),
     data = milk3,
     ylab = "Log(Liter mass)",
     pch = 18,
     cex =2,
     main = "Maternal Mass vs LItter mass duration")



#Lactation vs. gestation duration
plot(log(gest.month.NUM) ~ log(lacat.mo.NUM),
     data = milk3,
     ylab = "Log(Gestation dur)",
     pch = 18,
     cex =2,
     main = "Lactation vs. gestation duration")

```

* All of our predictors are highly correlated with each other
* "collinearity" of predictory or "multicollinearity" is major problem in regression





## Hard-code logs 

This will make the following code easier to do

```{r}
milk3$mass.fem.log <- log(milk3$mass.female)
milk3$lacat.mo.log <- log(milk3$lacat.mo.NUM)
milk3$gest.month.log <- log(milk3$gest.month.NUM)
milk3$mass.litter.log <- log(milk3$mass.litter)

```




# Pairs plots

* use pairs(..,lower.panel = NULL)
```{r, echo = F}
par(mfrow = c(1,1))
pairs(milk3[,c("fat.percent",
                    "mass.fem.log",
                    "lacat.mo.log",
                    "gest.month.log",
                    "mass.litter.log")],
      lower.panel = NULL)

```


# Add smoother & correlations

* see ?pairs and example using  panel.cor()

```{r, echo = F}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y))
    txt <- format(c(r, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, 
         txt#, 
         #cex = cex.cor * r
         )
}

pairs(milk3[,c("fat.percent",
                    "mass.fem.log",
                    "lacat.mo.log",
                    "gest.month.log",
                    "mass.litter.log")],
      lower.panel = panel.cor,
      col = 1,
      panel = panel.smooth)

```



